---
title: "Spark Histograms in Dataframe Columns"
date: '2017-09-11'
output:
html_document:
keep_md: yes
slug: spark-histograms-in-dataframe-columns
subtitle: ''
tags: []
categories: []
font-family: "DejaVu Sans"
mainfont: "DejaVu Sans"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F, comment ="")
```


A couple of weeks ago, I was looking for a package, I previously passed by that, prints summary statistics with inline histograms. I checked all my bookmarks and liked tweets, but I couldn't find it! So I asked on twitter. fortunately **[Maëlle Salmon](https://twitter.com/ma_salmon)** read the tweet and guided me to [`skimr`](https://github.com/ropenscilabs/skimr) by [ropenscilabs](https://github.com/ropenscilabs), who release many useful packages.


## skimr Summary Statistics

So basically `skimr` can show summary statistics, that one can go through quickly, handling different data types (numerics, factors, etc). For example, if we skim `storms`, and look at one variable  we can see the summaries as follows:

```{r load libraries}
# load libraries
library(tidyverse)
library(skimr)
library(knitr)

# Set locale (for windows)
slang <- Sys.setlocale("LC_CTYPE", "Chinese")
```

```{r skimr example}
# skim storms
storm_skim <- skim(storms)

storm_skim %>% 
  filter(var == "wind") %>% 
  kable()
```

So we can see the inline histogram, which might be useful while skimming a summary of many variables. And we can extract all the histograms for all the variables as follows:

```{r}
storm_skim %>% 
  filter(stat == "hist") %>% 
  kable()
```



## Spark Histograms Beyond Summary Statistics 

After exploring `skimr`, I thought what if I wanted to have spark histograms in a dataframe as a summary for a group?, i.e. something like ` df %>% group_by(x) %>% summarize(hist = plot_sark_hist())`. 

For the next example, I will use [WUZZUF Job Posts](https://www.kaggle.com/WUZZUF/wuzzuf-job-posts/downloads/Wuzzuf_Job_Posts_Sample.csv) dataset released on [Kaggle](https://www.kaggle.com/WUZZUF/wuzzuf-job-posts). 


```{r read dataset, echo=FALSE}
jobs <- read_csv("../../static/data/Wuzzuf_Job_Posts_Sample.csv")
```

```{r}
str(jobs)
```



```{r}
# m <- gapminder::gapminder %>% 
#   select(country, lifeExp) %>% 
#   group_by(country) %>% 
#   nest() %>% 
#   mutate(s = map(data, ~skimr::skim(.x) %>% filter(stat == "hist")) %>%
#            map("level") %>% 
#            map_chr(unlist))
```



```{r, results='asis'}
DT::datatable(head(storm_skim))
```



