---
title: 'Tidy Eval Meets ggplot2'  
subtitle: 'The Bang Bang Plots'
date: '2018-07-06'
slug: ggplot-tidyeval
tags: [R, ggplot2, tidyeval, tidyverse]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F, comment = "", eval = T)
```


Almost a year ago I wrote about my [My First Steps into The World of Tidy Eval](http://www.onceupondata.com/2017/08/12/my-first-steps-into-the-world-of-tidyeval/). At the end I [tweeted](https://twitter.com/OmaymaS_/status/887329478669729793) asking [Hadley Wickham](https://twitter.com/hadleywickham) and [Lionel Henrey](https://twitter.com/_lionelhenry) whether `ggplot2` was compatible with the `tidy eval`, They said that it was on the todo list.

Finally, `ggplot2 3.0.0` got [released](https://www.tidyverse.org/articles/2018/07/ggplot2-3-0-0/) last week with the support of `tidy eval`, so I thought it was time to write about it!

<blockquote class="twitter-tweet" data-lang="en"><p lang="en" dir="ltr">ggplot2 3.0.0 now on CRAN — <a href="https://t.co/m8fSDaaSbu">https://t.co/m8fSDaaSbu</a> 🎉 Headline features are tidy eval, sf support, position_dodge2(), &amp; viridis. Huge thanks to all 300+ contributors to this release, and particularly to <a href="https://twitter.com/ClausWilke?ref_src=twsrc%5Etfw">@ClausWilke</a>, <a href="https://twitter.com/kara_woo?ref_src=twsrc%5Etfw">@kara_woo</a>,  <a href="https://twitter.com/_lionelhenry?ref_src=twsrc%5Etfw">@_lionelhenry</a> and <a href="https://twitter.com/thomasp85?ref_src=twsrc%5Etfw">@thomasp85</a> <a href="https://t.co/igb9k0Z89C">pic.twitter.com/igb9k0Z89C</a></p>&mdash; Hadley Wickham (@hadleywickham) <a href="https://twitter.com/hadleywickham/status/1014511041609453568?ref_src=twsrc%5Etfw">July 4, 2018</a></blockquote>
<script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>


In this post, I will go over the traditional ways of using `ggplot2` then I will highlight some cases, in which `tidy eval` solves problems and provides more flexibility. 

## Dataset

In all the following examples We will use a simple dataset, which is `tips` from `reshape2` package. The dataset contains information about tips received by a waiter over a period of a few months. We will add a new column with `tip_percent` to use in our plots.

```{r load package}
## Load packages
library(reshape2)
library(tidyverse)
```
```{r print dataset}
## add a new column with tip_percent
tips <- tips %>% 
  mutate(tip_percent = tip/total_bill)

as_tibble(tips)
```


## ggplot2 the common way

### aes()

The common way of using `ggplot2` functions, and the first example most people try is using Non standard Evaluation (`NSE`), by passing variable names unquoted. For instance if you want to plot `tip_percent` versus `total_bill` you could simply write the following chunk of code.

```{r aes}
ggplot(tips, aes(x = total_bill,
                 y = tip_percent,
                 alpha = 0.5))+
  geom_point()+
  theme_minimal()+
  guides(alpha = FALSE)
```

### aes_string()

Another way is using `aes_string()` instead of `aes()`. In this case, you will  need to pass column names quoted as follows, which will give the same result.

```{r aes_string, eval = FALSE}
ggplot(tips, aes_string(x = "total_bill",
                 y = "tip_percent",
                 alpha = 0.5))+
  geom_point()+
  theme_minimal()+
  guides(alpha = FALSE)
```

This version could be helpful when you want to pass variables to `ggplot2` functions. For instance, you can handle a scenario, when you have pre-defined variables to use, or in case you take the variable names from a user through an interface. The following examples will work fine and give the same result.

```{r aes_string_var, eval = FALSE}

## define variables
x_var <- "total_bill"
y_var <- "tip_percent"
alpha_var <- 0.5

ggplot(tips, aes_string(x = x_var,
                 y = y_var,
                 alpha = alpha_var))+
  geom_point()+
  theme_minimal()+
  guides(alpha = FALSE)
```

And usually the use case for this, is the need to write a reusable function that wraps several `ggplot2` layers to work with whatever variables passed to it. For example, you can wrap the previous chunk in a function as follows

```{r aes_string_fn, eval = FALSE}
plot_points <- function(x, y, alpha){
  ggplot(tips, aes_string(x = x, y = y, alpha = alpha))+
  geom_point()+
  theme_minimal()+
  guides(alpha = FALSE)
}
```

Then you can either pass column names directly or the variable names.

```{r plot_points_ex1, eval = FALSE}
## pass column names directly
plot_points(x = "total_bill",
            y = "tip_percent",
            alpha = 0.5)

## pass variable names
plot_points(x = x_var,
            y = y_var,
            alpha = alpha_var)

```

And you can reuse the function to create similar plots with different variables. For instance `tip` on the y-axis instead of `tip_percent`

```{r plot_points_ex2, eval = FALSE}
## plot tip vs. total_bill
plot_points(x = "total_bill",
            y = "tip",
            alpha = 0.5)
```

## ggplot2 the tidy eval way

The previous options are good and valid in many cases, but there are some limitations. For instance, you cannot create a function and pass column names unquoted as follows. 

```{r define plot_points2, eval = FALSE}
plot_points2 <- function(x, y, alpha){
  ggplot(tips, aes(x = x, y = y, alpha = alpha))+
    geom_point()+
    theme_minimal()+
    guides(alpha = FALSE)
}

## invalid way
plot_points2(x = total_bill,
             y = tip_percent,
             alpha = 0.5)
```
This will just break and gives an error like:

`Error in FUN(X[[i]], ...) : object 'total_bill' not found`

Previously there was no straightforward way to handle this, but with `tidy eval` it became possible and easy.

### Unquoted column names / NSE 

You can use `enquo()` and the bang bang operator `!!` to create a function like `plot_points_tidyeval_01()` which takes the column names unquoted, and it will work properly.

```{r define plot_points_tidyeval_01}
plot_points_tidyeval_01 <- function(x, y, alpha){
  
  x_var <- enquo(x)
  y_var <- enquo(y)
  
  ggplot(tips, aes(x = !!x_var, y = !!y_var, alpha = alpha))+
    geom_point()+
    theme_minimal()+
    guides(alpha = FALSE)
}

## pass column names unquoted 
plot_points_tidyeval_01(x = total_bill,
                        y = tip_percent,
                        alpha = 0.5)
```

You might still see this as a simple use case that could be handled by `aes_string()`. But there are more things you can do, like variable facets.

### Variable Facets 

Suppose that you wanted to define a variable facet, you wouldn't have this flexibility even with `aes_string()`. But now you can implement it as follows. 

*Notice that `vars()` is a new helper inside `facet_grid` instead of the `~`.*


```{r define plot_points_tidyeval_02}
plot_points_tidyeval_02 <- function(x, y, alpha, facet){
  
  x_var <- enquo(x)
  y_var <- enquo(y)
  facet_var <- enquo(facet)
  
  ggplot(tips, aes(x = !!x_var, y = !!y_var, alpha = alpha))+
    geom_point()+
    theme_minimal()+
    guides(alpha = FALSE)+
    facet_grid(cols = vars(!!facet_var))}

```

So you can pass `sex` to the facet variable in the `plot_points_tidyeval_02()` function.

```{r facet_sex}

## use sex for facet
plot_points_tidyeval_02(x = total_bill,
                        y = tip_percent,
                        alpha = 0.5,
                        facet = sex)
```

or a different variable like `day`

```{r facet_day}
## use day for facet
plot_points_tidyeval_02(x = total_bill,
                        y = tip_percent,
                        alpha = 0.5,
                        facet = day)
```


You can even use the dots `...` to have more flexibility and allow the function to take more than one variable for faceting. Here you will need to use `quos()` to capture the dots and pass the result directly to the facet layer `facet_wrap()` as follows.


```{r facet_dots}
plot_points_tidyeval_03 <- function(x, y, alpha, ...){
  
  x_var <- enquo(x)
  y_var <- enquo(y)
  facet_vars <- quos(...)
  
  ggplot(tips, aes(x = !!x_var, y = !!y_var, alpha = alpha))+
    geom_point()+
    theme_minimal()+
    guides(alpha = FALSE)+
    facet_wrap(facet_vars, nrow = 2)
}

## use both sex and day for facets
plot_points_tidyeval_03(x = total_bill,
                        y = tip_percent,
                        alpha = 0.5,
                        sex, day)
```


## In Conclusion

**`tidy eval` with `ggplot2` offers new options and more capabilities. It solves problems people used to struggle with and gives a higher level of flexibilty in writing functions and programming. I just covered a couple of options in this post, but I am sure I will discover more stuff and hopefully write about them later.**

