---
title: yelpr Package for Yelp Fusion API
subtitle: 'Or My Story with Package Development'
date: '2018-04-08'
slug: yelpr
tags: [rstats, packages, yelp]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F, comment = "", out.width='100%')
```

Today I pushed a preliminary version of [yelpr](https://github.com/OmaymaS/yelpr) an R library for the [Yelp Fusion API](https://www.yelp.com/developers/documentation/v3/get_started), and my first public package. I am still working on it and intended to share it with others to get feedback. But I also thought about writing a blog post to tell my story with package development in general, and my thought process while developing `yelpr`.

## What was my journey with package development? 

When I started to learn R, end of 2015, I was benefiting from the great amount of packages developed by "others". However, at this time I didn't see myself among them, I was more into analysis per se. later on, mid 2016, I got to know more about the R community and all the collaboration behind many of the packages. I started to feel that I could contribute to package development but I thought that I needed to learn more before getting into this. In January 2017, I attended **Rstudio::conf** which was a great opportunity, where I got to see a representation of the R community. I came back positively charged feeling that regardless of my level of knowledge, I could be one of those "others".

I started to learn about package development through different resources, but the main source was [Hadley Wickham's](https://github.com/hadley) book [R Packages](http://r-pkgs.had.co.nz/). I also have to say that I definitely passed by [Hilary Parker's](https://github.com/hilaryparker) post [Writing an R package from scratch](https://hilaryparker.com/2014/04/29/writing-an-r-package-from-scratch/amp/), which I knew about from one of [NSSD Podcast](http://nssdeviations.com/) episodes.

However, during 2017, I just wrote packages for myself, as a way of organizing projects, or practicing. And because most of my work was confidential, I didn't have the chance to share my packages with others to get feedback, till I started to work on `yelpr`.

## So how did I think about yelpr?

In the previous couple of months, I was looking for different data sets to use for education purposes. I passed by different packages that enabled users to import data through APIs. For instance [bikedata](https://github.com/ropensci/bikedata) or [weathercan](https://github.com/ropensci/weathercan) by [rOpenSci](https://github.com/ropensci). I also appreciated all the work done to develop packages to deal with google cloud APIs or something I personally benefited from like [googlesheets](https://github.com/jennybc/googlesheets) by [Jenny Brian](https://github.com/jennybc). 

I felt all these packages are empowering, and I would start with a simple package to import data through an API. I had several ideas but either there were already developed packages or the APIs where not public. And I saw that Yelp had several official and unofficial [libraries](https://github.com/Yelp/yelp-api) in different languages for [Yelp Fusion API](https://www.yelp.com/developers/documentation/v3/get_started), but `R` was not on the list. So I thought about giving it a try.

## What does yelpr do?

`yelpr` is an interface to the **Business Endpoints** and **Events Endpoints** from [Yelp Fusion API](https://www.yelp.com/developers/documentation/v3/get_started). Using an API key, one could retrieve data about business and user reviews from millions of businesses across 32 countries.


### Business Endpoint 

The available functions to search businesses are:

- `business_search()`
- `business_lookup_id()`
- `business_search_phone()`
- `business_search_review()`
- `business_search_autocomplete()`
- `business_search_transaction()`
- `business_match_name()`


For example `business_search()` returns up to 1000 businesses based on the provided search criteria. The location OR (longitude+latitude) are required, plus Additional [optional parameters](https://www.yelp.com/developers/documentation/v3/business_search).

Here we can retrieve the first 5 results in 'New York' with the term 'chinese'

```{r example business_search, eval = FALSE}
# search businesses with keyword 'chinese' in 'New York'
business_ny <- business_search(api_key = key,
                location = 'New York',
                term = "chinese",
                limit = 5)
```


### Events Endpoint

The available functions to search events are:

- `event_lookup_id()`
- `event_search()`
- `event_search_featured()`

For example, `event_search_featured` returns the featured event, chosen by Yelp's community managers, for a given location.

So we can get details of featured event in the given location using the longitude and latitude.

```{r example event_search_featured, eval = FALSE}
# search featured event in the given location coordinates
event <- event_search_featured(api_key = key,
                      longitude = "-74.01385", latitude = "40.70387")
```


## What questions did I have in mind while developing yelpr?

So the current development version is the thing I decided to share for feedback and contributions. But let me mention some of the questions I had in mind while writing the functions and the documentation.

### Passing required and optional parameters

Since most of the functions have required and optional parameters, I was not sure whether to:

- build the function to take one list with both, or pass the required first then use `...` for the optional
- use `...` for the optional parameters or have them all with default values `NULL`

```{r parameters, eval = FALSE}

function (api_key,
          type = "lookup",
          parameters = list()) {
    # ......
}

## OR

function (api_key,
          type = "lookup",
          name, city, state, country, ...){
    # ......
}
```

And I ended up with the second choice.

### Handling errors

Since the functions deal with endpoints that return informative error messages, I thought whether to check for missing parameters inside each function and throw errors accordingly or to just make use of the errors returned in the response. 

I feel that the first approach adheres to the best practices in writing functions but in this version I just let the user see the errors in the repose.


### Test cases

I have been thinking about the test cases to write for these functions but I think I haven't practiced enough in this area. So I haven't pushed any tests yet.

### Reusability and minimal code

I was also thinking if any functions could be merged or inherit from the same class to reduce the repetition in code. But I started with separate functions to test them first and will dig deeper into this later.

### Documentation

Since some functions take up to 10 optional parameters, I wondered whether I should include their description in the documentation, or just refer to their link on the original API [endpoint link](https://www.yelp.com/developers/documentation/v3/business_search). For now I did a mix of both and same for the returned values.

## What's Next?

So these were my first steps and my thoughts on developing `yelpr` in specific. The previous practice on my private packages was beneficial that I could find my way using `devtools`. I could understand some of the confusing points like the usage of `@import` or the use of `Namespace`. But there remains some caveats that I expect to know more about when I receive feedback, get some reviews or discuss about the current version with all its flaws. 

**So if you have any feedback or suggestions, please comment or open an [issue](https://github.com/OmaymaS/yelpr/issues). I'd be glad to hear from you!**