---
title: Divergent Bars in ggplot2
author: ~
date: '2019-01-25'
slug: ggplot2-divergent-bars
categories: []
tags: [R, ggplot2]
subtitle: 'Step by step guide'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, error = FALSE, message = FALSE)
```


A couple of days ago I was asked by one of the participants I met at the [workshop](https://www.onceupondata.com/2019/01/04/datafest-tbilisi/) I gave in DataFest Tbilisi about a simple tutorial on plotting divergent bars in ggplot2 and other bar chart stuff. I was going to write a gist with explanation but I decided to write this post instead to share with others whenever they ask during/after a workshop or in other occasions.

In this post, I will give 2 step-by-step examples:

1st- normal bar chart *(with a diverging aspect)* using the `gapminder` dataset.

2nd- divergent bars using the `starwars` dataset from `dplyr` package.

## Example 1

### Dataset (gapminder)


```{r}
## load libraries
library(gapminder)
library(tidyverse)
```

Given the `gapminder` dataset which includes the the GDP per Capita for `r  n_distinct(gapminder$country)` countries, let's say we are interested in creating a plot to answers the question: **"What is the percentage change in GDP per capita between 1997 and 2007?"** . 

We need to have a dataframe with each country and the corresponding change, something like this:


```{r, echo = FALSE}
## create summarized data
gapminder_subset <- gapminder %>% 
  filter(year %in% c(1997, 2007),
         continent == "Americas") %>% 
  select(country, year, gdpPercap) %>% 
  spread(year, gdpPercap, sep = "_") %>% 
  mutate(gdp_change = 100*(year_2007-year_1997)/year_1997)

## display data
gapminder_subset
```

I will not explain the data manipulation steps and I will assume you have your data in this format.

### Data Visualization

#### Create basic bar plot

To give an answer to the question **"What is the percentage change in GDP per capita between 1997 and 2007?"**, you'd Probably start with putting the `country` on the x-axis and the `gdp_change` on the y_axis. 

Now you will have a couple of issues:

1- the countries names are overlapped.

2- the bars are ordered according to the `country` name, instead of the value of the `gdp_change` 


```{r bar_basic}
## plot gdp change versus country
ggplot(data = gapminder_subset,
       aes(x = country, y = gdp_change))+
  geom_bar(stat = "identity")
```

#### Flip axes

**For the first issue**, you could either rotate the country names or flip the whole chart to put the `country` on the y axis. We will go with the second solution, which could be achieved by one line `coord_flip()`.


```{r bar_flip}
## flip axes
ggplot(data = gapminder_subset,
       aes(x = country, y = gdp_change))+
  geom_bar(stat = "identity")+
  coord_flip()
```

#### Order countries by gdp_change

**For the second issue**, you can use `reorder()` to sort the bars based on a certain variable; which is `gdp_change` here.

```{r bar_reorder}
## order by value
ggplot(data = gapminder_subset,
       aes(x = reorder(country, gdp_change), y = gdp_change))+
  geom_bar(stat = "identity")+
  coord_flip()
```

#### Color bars based on `gdp_change` value

So with these few lines you have your basic bar chart. But you might want to color the bars based on whether the `gdp_change` is positive or negative. There is more than one way to do this.

**One fast way** is to pass an expression `gdp_change > 0` to the `fill` attribute. It is as if you say *"check whether the change is positive or negative and color the bars based on the result"*.

```{r bar_color}
ggplot(data = gapminder_subset,
       aes(x = reorder(country, gdp_change), y = gdp_change,
           fill = gdp_change > 0))+
  geom_bar(stat = "identity")+
  coord_flip()
```

**Another way** is to add a new column `gdp_change_positive` to `gapminder_subset` datfarame to hold this value. 

```{r}
gapminder_subset <- gapminder_subset %>%
  mutate(gdp_change_positive = gdp_change > 0)

head(gapminder_subset)  
```

Then you you can pass it to the `fill` attribute. This approach would be useful if you need to reuse the values in this column for other purposes like filtering or anything else. 

```{r bar_color2, eval = FALSE}
ggplot(data = gapminder_subset,
       aes(x = reorder(country, gdp_change), y = gdp_change,
           fill = gdp_change_positive))+
  geom_bar(stat = "identity")+
  coord_flip()
```

Now you can make some modifications to customize your figure. For instance:

- use `labs()` to define the axes labels and figure title.

- remove the guides about the fill colors by setting `fill == FALSE` in `guides()`

- and pick the theme you want to use.

#### Polish the final figure

```{r bar_color3}
ggplot(data = gapminder_subset,
       aes(x = reorder(country, gdp_change), y = gdp_change,
           fill = gdp_change_positive))+
  geom_bar(stat = "identity")+
  coord_flip()+
  labs(x = "Country", y = "GDP per Capita Change %",
       title = "Percentage chage in GDP per Capita",
       subtitles = "Americas (1997 to 2007)")+
  theme_minimal()+
  guides(fill = FALSE)
```

So this is a normal horizontal plot, but sometimes people consider a divergence aspect since the values can be higher or lower than base line *(here zero)*. But the real divergent bars usually have one more variable as shown in example 2.


## Example 2

### Dataset (starwars)

Let's say we have a summarized dataframe from the `starwars` dataset including the average height of the characters in each `homeworld` grouped by `gender`. You can visualize this in different ways according to what you want to emphasize. here we will focus on the divergent bars version.

```{r}
## summarize data
starwars_chars <- starwars %>%
  filter(gender %in% (c("male", "female"))) %>%
  filter(!is.na(homeworld)) %>%
  group_by(homeworld, gender) %>%
  summarise(average_height = median(height, na.rm = TRUE)) %>%
  group_by(homeworld) %>%
  # mutate(n = n()) %>%
  filter(n() == 2) %>%
  ungroup()

## display data
starwars_chars
```

### Data Visualization

If you want to have divergent bars, you need to have the values of one group as negatives. You can use `ifelse()` to multiply the values for the males by -1.

```{r}
starwars_chars <- starwars_chars %>%
  mutate(average_height = ifelse(gender == "female",
                                 average_height,
                                 -1*average_height))

```

Now you can simply plot a normal bar plot and use the fill color as the gender

```{r starwars1}
## plot divergent bars
starwars_chars %>%
  ggplot(aes(x = homeworld, y = average_height, fill = gender))+
  geom_bar(stat = "identity")+
  coord_flip()
```

You can notice one issue here in the x-axis values. you have negative heights which is not reasonable so you need to set these values manually to reflect the absolute values.

For this, you can use `scale_y_continuous()`. Notice that it is `scale_y_continuous()` not `scale_x_continuous()` because we deal with the original aesthetics we passed before flipping the coordinates. 

**But how to set the breaks and labels values?**

You have more than one option. For instance:

- you can pass a vector with the exact values.

- or you can use a more generic way that creates breaks based on the range of `average_height` in your data using `pretty()`. For example `pretty(starwars_chars$average_height)` will give the following values: 

```{r echo = FALSE}
pretty(starwars_chars$average_height)
```

We will use this to specify the `breaks`. And to make sure the displayed values are positive, we can pass the absolute values to the `labels` attribute.

```{r starwars2}
## calculate breaks values
breaks_values <- pretty(starwars_chars$average_height)

## create plot
starwars_chars %>%
  ggplot(aes(x = homeworld, y = average_height, fill = gender))+
  geom_bar(stat = "identity")+
  coord_flip()+
  scale_y_continuous(breaks = breaks_values,
                     labels = abs(breaks_values))+
  theme_minimal()
```
Now you can change the colors, add a vertical line at zero and customize your figure as you want.

```{r starwars3}
## create plot
starwars_chars %>%
  ggplot(aes(x = homeworld, y = average_height, fill = gender))+
  geom_hline(yintercept = 0)+
  geom_bar(stat = "identity")+
  coord_flip()+
  scale_y_continuous(breaks = breaks_values,
                     labels = abs(breaks_values))+
  theme_minimal()+
  scale_fill_manual(values = c("#bf812d", "#35978f"))
```

So these were simple examples with bar plots triggered by a question I received. But at the end picking a specific type of charts depends on the question in mind and the message one wants to deliver. And the good thing about `ggplot2` that it gives the cotrol over your plot to do whatever you want.