---
title: From H2O to POJO Models
subtitle: 'Getting started with a minimal example'
date: '2019-07-13'
slug: h2o-pojo-models
categories: []
tags: [R, h2o, java]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F, comment ="", eval = TRUE)
```

A while ago, I was experimenting with `h2o` and wanted to generate a Plain Old Java Object (POJO) model. I found the documentation useful but I decided to write a post with a simple example for future reference. 

In this post, we will see how to:

- build a simple `h2o` model in `R`.
- convert the model to `POJO`.
- create a `main` program in Java to use the `POJO` model. 
- compile and run the program.


### Initialize `h2o`

First of all, we need to load `h2o` library and connect to an `h2o` instance.

```{r, results=FALSE}
## load packages
library(h2o)

## initialize h2o
h2o.init()
```


### Import data

Once we make sure you are connected to h2o, we will import the `iris` dataset, which we are going to use in our example, using `h2o.importFile()`. 


```{r,  results=FALSE}
## import data
iris_path = system.file("extdata", "iris.csv", package = "h2o")

iris_hex = h2o.importFile(path = iris_path, destination_frame = "iris_hex")
```

We can check the returned dataframe, which has four columns corresponding to the flowers features and a the last column corresponds to the class/label.

```{r}
## display dataframe
iris_hex
```

### Build `h2o` model

For demo purposes, we will create a simple k-means model to cluster the entries in the `iris` dataset. 

```{r, results=FALSE}
## create kmeans model with 3 clusters
iris_km_model = h2o.kmeans(training_frame = iris_hex,
                           x = c("C1", "C2", "C3", "C4"),
                           k = 3,
                           model_id = "iris_km_model",
                           seed = 1000)
```

You can print some info about the model as follows:

```{r}
## print model summary
iris_km_model
```



### Convert `h2o` model to `POJO`

To convert your model to `POJO`, you can simply use `h2o.download_pojo()` with a path to save the java file.

```{r, eval = FALSE}
## download pojo model (replace "saved_models" with your path)
h2o.download_pojo(iris_km_model, path = here::here("saved_models"))
```

A java file, entitles `iris_km_model.java`, will be created in the given directory. 

*You can expand the following section to see the code corresponding to the `POJO` model in this file.*

<details>
  <summary><b>Click to expand</b></summary>
```java
/*
  Licensed under the Apache License, Version 2.0
    http://www.apache.org/licenses/LICENSE-2.0.html

  AUTOGENERATED BY H2O at 2019-07-13T17:22:02.421+02:00
  3.22.1.1
  
  Standalone prediction code with sample test data for KMeansModel named iris_km_model

  How to download, compile and execute:
      mkdir tmpdir
      cd tmpdir
      curl http:/localhost/127.0.0.1:54321/3/h2o-genmodel.jar > h2o-genmodel.jar
      curl http:/localhost/127.0.0.1:54321/3/Models.java/iris_km_model > iris_km_model.java
      javac -cp h2o-genmodel.jar -J-Xmx2g -J-XX:MaxPermSize=128m iris_km_model.java

     (Note:  Try java argument -XX:+PrintCompilation to show runtime JIT compiler behavior.)
*/
import java.util.Map;
import hex.genmodel.GenModel;
import hex.genmodel.annotations.ModelPojo;
import hex.genmodel.IClusteringModel;

@ModelPojo(name="iris_km_model", algorithm="kmeans")
public class iris_km_model extends GenModel implements IClusteringModel {
  public hex.ModelCategory getModelCategory() { return hex.ModelCategory.Clustering; }

  // Names of columns used by model.
  public static final String[] NAMES = NamesHolder_iris_km_model.VALUES;

  // Column domains. The last array contains domain of response column.
  public static final String[][] DOMAINS = new String[][] {
    /* C1 */ null,
    /* C2 */ null,
    /* C3 */ null,
    /* C4 */ null
  };

  public iris_km_model() { super(NAMES,DOMAINS,null); }
  public String getUUID() { return Long.toString(2946031392675382139L); }

  // Pass in data in a double[], pre-aligned to the Model's requirements.
  // Jam predictions into the preds[] array; preds[0] is reserved for the
  // main prediction (class for classifiers or value for regression),
  // and remaining columns hold a probability distribution for classifiers.
  public final double[] score0( double[] data, double[] preds ) {
    Kmeans_preprocessData(data,iris_km_model_MEANS.VALUES,iris_km_model_MULTS.VALUES,iris_km_model_MODES.VALUES);
    preds[0] = KMeans_closest(iris_km_model_CENTERS.VALUES, data, DOMAINS); 
    return preds;
  }

  // Pass in data in a double[], in a same way as to the score0 function.
  // Cluster distances will be stored into the distances[] array. Function
  // will return the closest cluster. This way the caller can avoid to call
  // score0(..) to retrieve the cluster where the data point belongs.
  public final int distances( double[] data, double[] distances ) {
    Kmeans_preprocessData(data,iris_km_model_MEANS.VALUES,iris_km_model_MULTS.VALUES,iris_km_model_MODES.VALUES);
    int cluster = KMeans_distances(iris_km_model_CENTERS.VALUES, data, DOMAINS, distances); 
    return cluster;
  }

  // Returns number of cluster used by this model.
  public final int getNumClusters() {
    int nclusters = iris_km_model_CENTERS.VALUES.length;
    return nclusters;
  }
}
// The class representing training column names
class NamesHolder_iris_km_model implements java.io.Serializable {
  public static final String[] VALUES = new String[4];
  static {
    NamesHolder_iris_km_model_0.fill(VALUES);
  }
  static final class NamesHolder_iris_km_model_0 implements java.io.Serializable {
    static final void fill(String[] sa) {
      sa[0] = "C1";
      sa[1] = "C2";
      sa[2] = "C3";
      sa[3] = "C4";
    }
  }
}
// Column means of training data
class iris_km_model_MEANS implements java.io.Serializable {
  public static final double[] VALUES = new double[4];
  static {
    iris_km_model_MEANS_0.fill(VALUES);
  }
  static final class iris_km_model_MEANS_0 implements java.io.Serializable {
    static final void fill(double[] sa) {
      sa[0] = 5.843333333333333;
      sa[1] = 3.053999999999999;
      sa[2] = 3.758666666666667;
      sa[3] = 1.1986666666666665;
    }
  }
}
// Reciprocal of column standard deviations of training data
class iris_km_model_MULTS implements java.io.Serializable {
  public static final double[] VALUES = new double[4];
  static {
    iris_km_model_MULTS_0.fill(VALUES);
  }
  static final class iris_km_model_MULTS_0 implements java.io.Serializable {
    static final void fill(double[] sa) {
      sa[0] = 1.2076330213409388;
      sa[1] = 2.3063033203973875;
      sa[2] = 0.5667583466456685;
      sa[3] = 1.3103399393571;
    }
  }
}
// Mode for categorical columns
class iris_km_model_MODES implements java.io.Serializable {
  public static final int[] VALUES = new int[4];
  static {
    iris_km_model_MODES_0.fill(VALUES);
  }
  static final class iris_km_model_MODES_0 implements java.io.Serializable {
    static final void fill(int[] sa) {
      sa[0] = -1;
      sa[1] = -1;
      sa[2] = -1;
      sa[3] = -1;
    }
  }
}
// Normalized cluster centers[K][features]
class iris_km_model_CENTERS implements java.io.Serializable {
  public static final double[][] VALUES = new double[3][];
  static {
    iris_km_model_CENTERS_0.fill(VALUES);
  }
  static class iris_km_model_CENTERS_0_0 implements java.io.Serializable {
    public static final double[] VALUES = new double[4];
    static {
      iris_km_model_CENTERS_0_0_0.fill(VALUES);
    }
    static final class iris_km_model_CENTERS_0_0_0 implements java.io.Serializable {
      static final void fill(double[] sa) {
        sa[0] = -0.0685873626223117;
        sa[1] = -0.8873945545098232;
        sa[2] = 0.3438624614956358;
        sa[3] = 0.2839741837806723;
      }
    }
}
  static class iris_km_model_CENTERS_0_1 implements java.io.Serializable {
    public static final double[] VALUES = new double[4];
    static {
      iris_km_model_CENTERS_0_1_0.fill(VALUES);
    }
    static final class iris_km_model_CENTERS_0_1_0 implements java.io.Serializable {
      static final void fill(double[] sa) {
        sa[0] = 1.1276273336771014;
        sa[1] = 0.08687075840163734;
        sa[2] = 0.9821922147369432;
        sa[3] = 0.9954215739316106;
      }
    }
}
  static class iris_km_model_CENTERS_0_2 implements java.io.Serializable {
    public static final double[] VALUES = new double[4];
    static {
      iris_km_model_CENTERS_0_2_0.fill(VALUES);
    }
    static final class iris_km_model_CENTERS_0_2_0 implements java.io.Serializable {
      static final void fill(double[] sa) {
        sa[0] = -1.0111913832028125;
        sa[1] = 0.8394944086246519;
        sa[2] = -1.3005214861029282;
        sa[3] = -1.2509378621062437;
      }
    }
}
  static final class iris_km_model_CENTERS_0 implements java.io.Serializable {
    static final void fill(double[][] sa) {
      sa[0] = iris_km_model_CENTERS_0_0.VALUES;
      sa[1] = iris_km_model_CENTERS_0_1.VALUES;
      sa[2] = iris_km_model_CENTERS_0_2.VALUES;
    }
  }
}
```  
  
  
</details>  


### Create `main` java program

Now you need to write the main program `main.java` and save it with in the same directory with `iris_km_model.java`. The program is supposed to take four values and return the predicted cluster. 

Note that you need to:

-  provide the the pojo model name included in the file downloaded earlier in the previous step(`iris_km_model`).

- use the right class for your model, which is `ClusteringModelPrediction` here. You can see further details about classes on the [POJO Model Javadoc page](http://h2o-release.s3.amazonaws.com/h2o/master/3591/docs-website/h2o-genmodel/javadoc/index.html)



```java
import java.io.*;
import hex.genmodel.easy.RowData;
import hex.genmodel.easy.EasyPredictModelWrapper;
import hex.genmodel.easy.prediction.*;

public class main {
  private static String modelClassName = "iris_km_model";

  public static void main(String[] args) throws Exception {
    hex.genmodel.GenModel rawModel;
    rawModel = (hex.genmodel.GenModel) Class.forName(modelClassName).newInstance();
    EasyPredictModelWrapper model = new EasyPredictModelWrapper(rawModel);

    RowData row = new RowData();
    row.put("C1", args[0]);
    row.put("C2", args[1]);
    row.put("C3", args[2]);
    row.put("C4", args[3]);

    ClusteringModelPrediction p = model.predictClustering(row);

    System.out.printf("Input values: %s %s %s %s \n", args[0], args[1], args[2], args[3]);
    System.out.printf("cluster: %s", p.cluster);
  }
}
```


### Compile java program 

Now it is time to compile your program, but first you need to download `h2o-genmodel.jar` *in the same directory* with `iris_km_model.java` and `main.java`. 

```
$ cd saved_models
$ curl http://localhost:54321/3/h2o-genmodel.jar > h2o-genmodel.jar
```

Having the three files, you can compile your program as follows:

```
$ javac -cp h2o-genmodel.jar -J-Xmx2g -J-XX:MaxPermSize=128m iris_km_model.java main.java
```

If things work fine you will get no errors and you will find new `.class` files generated in the same directory.

### Make predictions

Now you are ready to use the compiled program and make predictions. For this, you should write `java -cp ".;h2o-genmodel.jar" main` followed by the four expected inputs *(corresponding to C1, C2, C3, C4)* as follows:

```
$java -cp ".;h2o-genmodel.jar" main  6.4 3.2 5.3 2.3
```

The printed result will include whatever format you specified in the `main.java` program. Here I set it to return the input values then the predicted cluster.

```
Input values: 6.4 3.2 5.3 2.3
cluster: 1
```

**NOTE**:

- I used `;` in `$java -cp ".;h2o-genmodel.jar"` as I am using Windows. Probably it will differ with other operating systems and you will need to use `:`.

- There could be better ways to do the same thing, but this was a minimal example with some basics for demo purposes. 


### Session info

```
> sessionInfo()
R version 3.5.0 (2018-04-23)
Platform: x86_64-w64-mingw32/x64 (64-bit)
Running under: Windows >= 8 x64 (build 9200)

Matrix products: default

locale:
[1] LC_COLLATE=English_United States.1252  LC_CTYPE=English_United States.1252   
[3] LC_MONETARY=English_United States.1252 LC_NUMERIC=C                          
[5] LC_TIME=English_United States.1252    

attached base packages:
[1] stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
[1] h2o_3.22.1.1

loaded via a namespace (and not attached):
 [1] compiler_3.5.0  bookdown_0.9    htmltools_0.3.6 tools_3.5.0     RCurl_1.95-4.12 yaml_2.2.0     
 [7] Rcpp_1.0.1      rmarkdown_1.12  blogdown_0.11   knitr_1.22      jsonlite_1.6    xfun_0.5       
[13] digest_0.6.18   bitops_1.0-6    evaluate_0.13 
```


## Extra Resources 


- [POJO Quick Start](https://h2o-release.s3.amazonaws.com/h2o/rel-turing/1/docs-website/h2o-docs/pojo-quick-start.html#what-is-a-pojo)

- [POJO Model Javadoc page](http://h2o-release.s3.amazonaws.com/h2o/master/3591/docs-website/h2o-genmodel/javadoc/index.html)



